---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then
     regenerate with devtools::build_readme(). Do not use knitr::knit(): it
     processes the code but leaves this YAML header in the output as literal
     text, which GitHub and pkgdown both render verbatim. -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/statmodels7/covstructs7/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/statmodels7/covstructs7/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/statmodels7/covstructs7/graph/badge.svg)](https://app.codecov.io/gh/statmodels7/covstructs7)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(covstructs7)
```

# covstructs7

Every R package that fits a model with a covariance matrix carries its own
parametrisation of it, written inside the routine that needs it. `nlme` has
the `pdMat` classes, `lme4` has its `theta` vector, `glmmTMB` has its
covariance structures — three private rewritings of the same mathematics, none
exposing a derivative, none reusable from outside.

A covariance structure is a fixed mathematical object. The map from an
unconstrained vector to a matrix in some constrained set, its inverse, its
derivatives, and the quantities a likelihood asks of the matrix are all
determined once and for all, so they can be written once and built on.

`{covstructs7}` makes a structure an object that can be **evaluated,
inverted, differentiated to second order, and asked for its log-determinant
and its solves** — including when the matrix is **rank deficient**, which a
covariance may not be but a precision routinely is.

It is the matrix-parameter layer of [statmodels7](https://statmodels7.github.io),
an S7 toolkit for statistical modelling, alongside
[linkfunctions7](https://statmodels7.github.io/linkfunctions7/),
[distributions7](https://statmodels7.github.io/distributions7/),
[optimizers7](https://statmodels7.github.io/optimizers7/) and
[basis7](https://statmodels7.github.io/basis7/).

## Installation

```{r eval=FALSE}
# install.packages("pak")
pak::pak("statmodels7/covstructs7")
```

## A structure is an object

```{r}
s <- log_cholesky(3)
s

eta <- c(0.1, -0.2, 0.3, 0.5, -0.4, 0.2)
round(struct_matrix(s, eta), 4)
```

The map is a bijection onto the positive definite cone, so it inverts exactly:

```{r}
max(abs(struct_free(s, struct_matrix(s, eta)) - eta))
```

## Not a link, and why

A link in `{linkfunctions7}` is a **scalar** bijection with a diagonal
Jacobian, and everything built on it depends on that diagonality. A map from
$\mathbb{R}^d$ to a matrix has a full Jacobian, so it is a different contract
and a different class. The two **compose**: the diagonal block of a structured
matrix is a scalar map, and those are `{linkfunctions7}` links reused as they
are.

```{r}
struct_matrix(diag_struct(3, link = linkfunctions7::sqrt_link()), c(1, 2, 3))
```

## Rank-deficient precisions are the point

A spline penalty is singular by construction: its null space is the functions
it leaves alone. Reading it as a prior makes that prior improper, which is
exactly what makes it a legitimate **penalty** and not a legitimate density.

```{r}
p <- crossprod(diff(diag(6), differences = 2))
s <- scaled_struct(p)
s
```

The null space is the constants and the straight lines — the functions a
second-difference penalty leaves alone — and the structure carries an
orthonormal basis for it:

```{r}
lin <- cbind(1, 1:6)
c(penalty_annihilates_them = max(abs(p %*% lin)),
  null_basis_spans_the_same = max(abs(qr.resid(qr(lin), s@null_basis))))
```

The log-determinant becomes the log **pseudo**-determinant, and its derivative
is the rank — a constant, whatever the scale:

```{r}
c(at_small_scale = unname(struct_dlogdet(s, -6)),
  at_large_scale = unname(struct_dlogdet(s, 6)),
  rank = s@rank)
```

That derivative is what makes the scale estimable at all. Writing a penalty as
a negative log prior, $\frac{\lambda}{2}\beta^\top P\beta -
\frac{r}{2}\log\lambda$, the stationary point is $\lambda = r/(\beta^\top P
\beta)$; dropping the constant leaves a derivative of one sign and sends
$\lambda$ to zero.

```{r}
set.seed(1)
beta <- rnorm(6)
q <- drop(t(beta) %*% p %*% beta)

c(closed_form = s@rank / q,
  numerical = optimize(function(l) l / 2 * q - s@rank / 2 * log(l),
                       c(1e-8, 1e8), tol = 1e-12)$minimum)
```

## Rank comes from the components, never from an assembled matrix

The mathematical rank of $\lambda_1 P_1 + \lambda_2 P_2$ does not depend on
the positive scalars. A count of eigenvalues above a relative tolerance does,
and a fitted model with smoothing parameters far apart is ordinary:

```{r}
p1 <- kronecker(crossprod(diff(diag(8), differences = 2)), diag(4))
p2 <- kronecker(diag(8), crossprod(diff(diag(4), differences = 2)))

count_ev <- function(m) {
  e <- svd(m, nu = 0, nv = 0)$d
  sum(e > 1e-8 * max(e))
}

c(balanced = count_ev(p1 + p2),
  ratio_1e12 = count_ev(1e-6 * p1 + 1e6 * p2),
  from_components = struct_null_basis(list(p1, p2))$rank)
```

So `struct_null_basis()` stacks the individually normalised components — the
null space of a sum of positive semidefinite matrices is the intersection of
their null spaces — and membership is afterwards tested against that basis.

## A user-defined structure needs only its map

Everything else has a numerical method registered on the base class, so a new
structure is a subclass and one method.

```{r}
Ar1 <- S7::new_class("Ar1", parent = covstruct)

S7::method(struct_matrix, Ar1) <- function(s, eta, ...) {
  p <- s@dimension
  exp(eta[1]) * tanh(eta[2])^abs(outer(seq_len(p), seq_len(p), "-"))
}

ar1 <- Ar1(
  struct_name = "ar1", dimension = 4L, n_free = 2L,
  free_names = c("scale", "rho"), rank = 4L,
  null_basis = matrix(numeric(0), 4, 0), role = "covariance",
  struct_params = list()
)

# never implemented, yet available
round(struct_dmatrix(ar1, c(0.2, 0.6))[["rho"]], 4)
struct_logdet(ar1, c(0.2, 0.6))
```

## Validating a structure

`check_covstruct()` runs nine checks, each against a route the implementation
does not take. A quantity that comes from the base class is reported as
**not checked**, not as passed: comparing a finite difference with a finite
difference is the same arithmetic twice.

```{r}
invisible(check_covstruct(log_cholesky(3)))
invisible(check_covstruct(ar1))
```

## What is in the box

| | |
|---|---|
| families | `log_cholesky()`, `diag_struct()`, `scalar_struct()`, `scaled_struct()` |
| maps | `struct_matrix()`, `struct_free()` |
| derivatives | `struct_dmatrix()`, `struct_d2matrix()` |
| likelihood pieces | `struct_logdet()`, `struct_dlogdet()`, `struct_d2logdet()`, `struct_solve()`, `struct_factor()` |
| tools | `check_covstruct()`, `struct_is_numerical()`, `struct_null_basis()`, `print()` |

## Related

- [linkfunctions7](https://statmodels7.github.io/linkfunctions7/) — link
  functions with exact derivatives to fourth order
- [distributions7](https://statmodels7.github.io/distributions7/) —
  distributions carrying exact derivatives of the log-likelihood
- [optimizers7](https://statmodels7.github.io/optimizers7/) — optimisation
  algorithms and stopping rules as objects
- [basis7](https://statmodels7.github.io/basis7/) — basis expansions with
  exact derivatives, integrals and Gram matrices
- [the book](https://statmodels7.github.io/book/) — the mathematics behind
  the toolkit
